The algebraic hierarchy in type theory using type classes

### 2010-05-28 defManSSR.tex
Use Unicode in listings.

% Vernacular commands
morekeywords=[1]{
Section, Module, End, Require, Import, Export,
Variable, Variables, Parameter, Parameters, Axiom, Hypothesis, Hypotheses,
Notation, Local, Tactic, Reserved, Scope, Open, Close, Bind, Delimit,
Definition, Let, Ltac, Fixpoint, CoFixpoint, Add, Morphism, Relation,
Implicit, Arguments, Set, Unset, Contextual, Strict, Prenex, Implicits,
Inductive, CoInductive, Record, Structure, Canonical, Coercion,
Theorem, Lemma, Corollary, Proposition, Fact, Remark, Example,
Proof, Goal, Save, Qed, Defined, Hint, Resolve, Rewrite, View,
Search, Show, Print, Printing, All, Graph, Projections, inside, outside},

@@

% Enables ASCII chars 128 to 255
extendedchars=true,

>>>>

% Vernacular commands
morekeywords=[1]{
Section, Module, End, Require, Import, Export,
Variable, Variables, Parameter, Parameters, Axiom, Hypothesis, Hypotheses,
Notation, Local, Tactic, Reserved, Scope, Open, Close, Bind, Delimit,
Definition, Let, Ltac, Fixpoint, CoFixpoint, Add, Morphism, Relation,
Implicit, Arguments, Set, Unset, Contextual, Strict, Prenex, Implicits,
Inductive, CoInductive, Record, Structure, Canonical, Coercion,
Context, Class, Global, Instance, Program, Infix,
Theorem, Lemma, Corollary, Proposition, Fact, Remark, Example,
Proof, Goal, Save, Qed, Defined, Hint, Resolve, Rewrite, View,
Search, Show, Print, Printing, All, Graph, Projections, inside, outside},

@@

% Enables ASCII chars 128 to 255
extendedchars=false,

### 2010-05-28 jar.tex
Use Unicode in listings.

\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{listings}
% The package listingsutf8 supports utf8 for \lstinputlisting
% However, we do not want to put all our code snippets in separate files.
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
\usepackage{color}
\definecolor{dkblue}{rgb}{0,0.1,0.5}
\definecolor{lightblue}{rgb}{0,0.5,0.5}
\definecolor{dkgreen}{rgb}{0,0.4,0}
\definecolor{dk2green}{rgb}{0.4,0,0}
\definecolor{dkviolet}{rgb}{0.6,0,0.8}
\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum. To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
\end{abstract}

@@

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we
might define:
% \lstinputlisting[language=SSR,inputencoding=utf8/latin1]{ReflexiveRelation.v}

\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: $\Pi$ x, rr_rel x x }.
\end{lstlisting}

@@

As a first step we would define \lstinline|ReflexiveRelation| as:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: relation rr_carrier
    ; rr_proof: $\Pi$ x, rr_rel x x }.
\end{lstlisting}

@@

A natural solution is to pull the relation out of \lstinline|ReflexiveRelation| (and its siblings) as well:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
    { rr_proof: $\Pi$ x, rr_rel x x }.
\end{lstlisting}

or equivalently:
\begin{lstlisting}
  Definition ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
    $\Pi$ x, rr_rel x x.
\end{lstlisting}

@@

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of \lstinline|Monoid| that bundles the
relation. Then if we proceed to define \lstinline|SemiRing| as a double \lstinline|Monoid|, we face a dilemma:
\begin{lstlisting}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: $\Pi$ x: A, H
\end{lstlisting}

@@

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{lstlisting}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: forall x, 0 * x == 0 }.
\end{lstlisting}

@@

% theory/ua_subalgebraT.v
% find . -iname "*.v" |xargs grep "niverse"
We define congruence:
\begin{lstlisting}
Class Congruence: Prop :=
    { congruence_proper:> $\Pi$ s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{lstlisting}

@@

\begin{lstlisting}
Class Category O`{!Arrows O}`{$\Pi$ x y:O, Equiv(x $\to$ y)}
                `{!CatId O}`{!CatComp O}:Prop :=
  { arrow_equiv:> $\Pi$ x y, Setoid (x $\to$ y)
  ; comp_proper:> $\Pi$ x y z,
    Proper (equiv $\Rightarrow$ equiv $\Rightarrow$ equiv)%signature
    (comp: (y $\to$ z) $\to$ (x$\to$ y) $\to$ x $\to$ z)
  ; comp_assoc w x y z (a: w $\to$ x) (b: x $\to$ y) (c: y $\to$ z):
      c $\circ$ (b $\circ$ a) = (c $\circ$ b) $\circ$ a
  ; id_l `(a: x $\to$ y): cat_id $\circ$ a = a
  ; id_r `(a: x $\to$ y): a $\circ$ cat_id = a }.
\end{lstlisting}

@@

%theory/categories.v
Natural transformations are formalized as:
\begin{lstlisting}
   Context `{Category C} `{Category D}
   `{!Functor (F: C $\to$ D) Fa} `{!Functor (G: C $\to$ D) Ga}.
   Class NaturalTransformation ($\eta$: $\Pi$ c, F c $\to$ G c): Prop :=
     natural: $\Pi$ (x y: C) (f: x $\to$ y), $\eta$ y $\circ$ fmap F f = fmap G f $\circ \eta$ x.
\end{lstlisting}

@@

 An expression is only meaningful in the context of a variable
assignment:\footnote{Categorically: \lstinline|nat| is
an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
\begin{lstlisting}
Definition Value := nat.
Definition Vars V := V -> Value.

@@

%Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
%$TX:=1+T^2X$, may be seen as the prototypical monad. 
The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad\footnote{on the category of \lstinline|Type|s with extensional functions.%\marginpar{Proof in classquote}
}, provides a bind operation which describes the arrows in the Kleisli category. Concretely,
\begin{lstlisting}
Fixpoint bind {V W} (f: V -> Expr W) (e: Expr V): Expr W :=
  match e with
  | Zero => Zero
  | Mult a b => Mult (bind f a) (bind f b)
  | Var v => vs v
  end.
\end{lstlisting}

@@

\noindent We define some simple combinators for variable packs:
%
\begin{lstlisting}
Definition novars: Vars False := False_rect _.
Definition singlevar (x: Value): Vars unit := fun _ => x.
Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
  fun i => if i then a j else b j.
\end{lstlisting}

@@

\begin{lstlisting}
Definition map_var {V W: Type} (f: V -> W):
    Expr V -> Expr W :=
  fix F (e: Expr V): Expr W :=
    match e with
    | Mult a b => Mult (F a) (F b)
    | Zero => Zero
    | Var v => Var (f v)
    end.
\end{lstlisting}

@@

We have defined a light-weight quoting function. The following code turns the goal into \lstinline|(eval variable_pack quote)|.
Here \lstinline|eval_quote'| is a modification of \lstinline|eval_quote| which works in the empty context:
\begin{lstlisting}
Goal forall x y (P: Value $\to$ Prop), P ((x * y) * (x * 0)).
  intros. rewrite <- (eval_quote' _).
\end{lstlisting}

>>>>

\documentclass[a4paper,10pt,runningheads]{llncs}
\usepackage{url}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{upgreek}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenx}
\usepackage{comment}
\newcommand{\N}{\ensuremath{\mathbbm{N}}}
\newcommand{\Z}{\ensuremath{\mathbbm{Z}}}
\newcommand{\Q}{\ensuremath{\mathbbm{Q}}}
\usepackage{color}

\definecolor{dkblue}{rgb}{0,0.1,0.5}
\definecolor{lightblue}{rgb}{0,0.5,0.5}
\definecolor{dkgreen}{rgb}{0,0.4,0}
\definecolor{dk2green}{rgb}{0.4,0,0}
\definecolor{dkviolet}{rgb}{0.6,0,0.8}

% listings:

\usepackage{listings}
  % The package listingsutf8 supports utf8 for \lstinputlisting
  % However, we do not want to put all our code snippets in separate files.

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi\ $}}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
}

\begin{document}
\title{The algebraic hierarchy in type theory using type classes}
\author{Bas Spitters \and Eelis van der Weegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
We present a new formalization of the algebraic hierarchy in Coq, exploiting its new type class mechanism to make practical a solution formerly thought infeasible. Our approach addresses both traditional challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a bare minimum. To support mathematically sound abstract interfaces for $\N$, $\Z$, and $\Q$, our formalization includes portions of category theory and multisorted universal algebra.
Algebra flourishes by the interplay between syntax and semantics. The Prolog-like
abilities of type class unification allow us to conveniently define a quote function thus facilitating the use of reflective techniques.
\end{abstract}

@@

How should we define reflexive, transitive, and symmetric relations? In a first attempt, we might define:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}

@@

As a first step we would define \lstinline|ReflexiveRelation| as:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type): Type :=
    { rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}

@@

A natural solution is to pull the relation out of \lstinline|ReflexiveRelation| (and its siblings) as well:
\begin{lstlisting}
  Record ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
    { rr_proof: Π x, rr_rel x x }.
\end{lstlisting}

or equivalently:
\begin{lstlisting}
  Definition ReflexiveRelation (carrier: Type) (rel: relation carrier): Prop :=
    Π x, rr_rel x x.
\end{lstlisting}

@@

At the risk of overstressing the point, let us look at how the same problem and solution keep
appearing in composite structures. Suppose we use a definition of \lstinline|Monoid| that bundles the
relation. Then if we proceed to define \lstinline|SemiRing| as a double \lstinline|Monoid|, we face a dilemma:
\begin{lstlisting}
  Record SemiRing A: Type :=
    { mult_monoid: Monoid A 
    ; plus_monoid: Monoid A 
    ; plus_comm: Commutative (mon_op plus_monoid)
    ; mult_comm: Commutative (mon_op mult_monoid)
    ; mult_0_l: Π x: A, H
\end{lstlisting}

@@

\section{Putting it all together}\label{classes}
Combining the ideas in the previous sections we end up with a formalization of algebraic
structures where we use the structure as (implicit) parameters, but
pack the proofs. This can be nicely organized by using canonical names. For instance, we define:
\begin{lstlisting}
Class SemiRing A {e: Equiv A}{plus: RingPlus A}{mult: RingMult A}
                 {zero: RingZero A}{one: RingOne A}:Prop :=
  { semiring_mult_monoid:> Monoid A (op:=mult)(unit:=one)
  ; semiring_plus_monoid:> Monoid A (op:=plus)(unit:=zero)
  ; semiring_plus_comm:> Commutative plus
  ; semiring_mult_comm:> Commutative mult
  ; semiring_distr:> Distribute mult plus
  ; mult_0_l: Π x, 0 * x = 0 }.
\end{lstlisting}

@@

% theory/ua_subalgebraT.v
% find . -iname "*.v" |xargs grep "niverse"
We define congruence:
\begin{lstlisting}
Class Congruence: Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _
    }.
\end{lstlisting}

@@

\begin{lstlisting}
Class Category O `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
    `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z,
    Proper (equiv ==> equiv ==> equiv) comp
  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{lstlisting}

@@

%theory/categories.v
Natural transformations are formalized as:
\begin{lstlisting}
   Context `{Category C} `{Category D}
   `{!Functor (F: C ⟶ D) Fa} `{!Functor (G: C ⟶ D) Ga}.
   Class NaturalTransformation (η: Π c, F c ⟶ G c): Prop :=
     natural: Π (x y: C) (f: x ⟶ y), η$$ y ∘ fmap F f = fmap G f ∘ η$$ x.
\end{lstlisting}
  % Todo: Having to add "$$" to get whitespace is awful. we can't add "\ " in η's literate-replacement though, because then we also get space before the colon in "η: t".

@@

 An expression is only meaningful in the context of a variable
assignment:\footnote{Categorically: \lstinline|nat| is
an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
\begin{lstlisting}
Definition Value := nat.
Definition Vars V := V → Value.

@@

%Monads are trees with grafting~\cite{MonadsGrafting}: the tree monad
%$TX:=1+T^2X$, may be seen as the prototypical monad. 
The \lstinline|Expr| monad, i.e.\ the free \lstinline|Expr|-algebra monad\footnote{on the category of \lstinline|Type|s with extensional functions.%\marginpar{Proof in classquote}
}, provides a bind operation which describes the arrows in the Kleisli category. Concretely,
\begin{lstlisting}
Fixpoint bind {V W} (f: V → Expr W) (e: Expr V): Expr W :=
  match e with
  | Zero => Zero
  | Mult a b => Mult (bind f a) (bind f b)
  | Var v => vs v
  end.
\end{lstlisting}

@@

\noindent We define some simple combinators for variable packs:
%
\begin{lstlisting}
Definition novars: Vars False := False_rect _.
Definition singlevar (x: Value): Vars unit := λ _ => x.
Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
  λ i => if i then a j else b j.
\end{lstlisting}

@@

\begin{lstlisting}
Definition map_var {V W: Type} (f: V → W):
    Expr V → Expr W :=
  fix F (e: Expr V): Expr W :=
    match e with
    | Mult a b => Mult (F a) (F b)
    | Zero => Zero
    | Var v => Var (f v)
    end.
\end{lstlisting}

@@

We have defined a light-weight quoting function. The following code turns the goal into \lstinline|(eval variable_pack quote)|.
Here \lstinline|eval_quote'| is a modification of \lstinline|eval_quote| which works in the empty context:
\begin{lstlisting}
Goal Π x y (P: Value $\to$ Prop), P ((x * y) * (x * 0)).
  intros. rewrite <- (eval_quote' _).
\end{lstlisting}
